\textsc{HubbleCLIP}: Associating Astronomical Observations and Natural Language with Multi-Modal Models

### 2024-01-14 paper/main.tex
Described results with single-concept summaries

The top 3 text associations are shown for each image query. The `ground truth' summarized abstract is shown in the right column. The base model is seen to return a mix of relevant and less-relevant associations. While it can often return the nature of objects imaged, we observe it to seldom return scientific phenomena (e.g., `dark matter' as successfully done by the fine-tuned model in the second row). The third row (supernova 1987A) highlights interesting behavior -- the base model erroneously attributes the object at the center of the image to a gravitational lens or protoplanetary disk, while the fine-tuned model correctly identifies it as a supernova remnant (which play a crucial role for interstellar chemistry -- another returned snippet).

@@

>>>>

The top 3 text associations are shown for each image query. The `ground truth' summarized abstract is shown in the right column. The base model is seen to return a mix of relevant and less-relevant associations. While it can often return the nature of objects imaged, we observe it to seldom return scientific phenomena (e.g., `dark matter' as successfully done by the fine-tuned model in the second row). The third row (supernova 1987A) highlights interesting behavior -- the base model erroneously attributes the object at the center of the image to a gravitational lens or protoplanetary disk, while the fine-tuned model correctly identifies it as a supernova remnant (which play a crucial role for interstellar chemistry -- another retrieved text snippet).

\paragraph*{Cosine similarity distribution}

The top row of Fig.~\ref{fig:sim_valtrain} shows the distribution of cosine similarities between corresponding image and text embeddings, $x_i$ and $y_i$, for the LLM-summarized abstracts using the fine-tuned CLIP model (red line), and for the base CLIP model (blue line), shown for the set of validation pairs (top left) and same number of training set pairs (top right). Distributions shuffling the order of text embeddings, randomizing the image-text correspondence, is shown as dashed lines. The distribution for the base model is seen to be sharply peaked at a specific value, showing little diversity and being very similar between the shuffled (dashed blue) and non-shuffled (solid blue) versions. Histograms for the fine-tuned model show a clear separation between the shuffled and corresponding text-image pair versions.

Distributions for a set of training samples (top right) are visible more peaked towards higher values compared to those evaluated on validation samples (top left), indicating potential distribution shift and/or a degree of overfitting.

Finally, in Fig.~\ref{fig:sim_valtrain} (bottom) we show the distribution of corresponding image-text pair cosine similarities for the single-concept summaries, evaluated using the CLIP model fine-tuned on summarized abstracts (red) and on the single-concept summaries themselves (purple). These distributions are peaked towards only marginally higher values compared to the baseline shuffled version, indicating that assigning $\mathcal O(100)$ labels to the images results in a relatively weak association signal compared to using noisy captions.

@@

\begin{figure*}[!h]
\includegraphics[width=0.95\textwidth]{plots/sim_valtrain.pdf}
\centering\includegraphics[width=0.45\textwidth]{plots/sim_summ1.pdf}
\caption{Distribution of cosine similarities between corresponding image and text embeddings, $x_i$ and $y_i$. (Top row) For the LLM-summarized abstracts using the fine-tuned CLIP model (red line), and for the base CLIP model (blue line), shown for (top left) the set of validation pairs and (top right) same number of training set pairs. Versions with the ordering of text embeddings shuffled are shown in dashed. (Bottom) For the single-concept summaries, when evaluated using the CLIP model fine-tuned on LLM-summarized abstracts (red) and fine-tuned on the single-concept summaries themselves (purple).}
\label{fig:sim_valtrain}
\end{figure*}
